---
title: "Declaring and enforcing an architecture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Declaring and enforcing an architecture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(aRchtest)
```

An architectural agreement that lives in a README is not enforced. This vignette
shows how to turn one into a test that fails the moment someone crosses the
boundary.

The examples run against a tiny layered project shipped with the package:

```{r}
project <- system.file("examples", "layered", package = "aRchtest")

list.files(project, recursive = TRUE)
```

It has three layers. `R/ui_dashboard.R` renders things, `R/domain_pricing.R`
holds business rules, and `R/infra_store.R` reads and writes files.

## A rule for a package project

A rule names an agreement, selects the modules it governs, and states a
constraint. Nothing happens until you check it -- a rule is inert data.

```{r}
ui_layer <- rule(
  "UI must not compute statistics",
  why = "Presentation stays thin so the domain owns the numbers"
) |>
  modules_matching("R/ui_*.R") |>
  must_not_call(packages = "stats")

ui_layer
```

`arch_check()` runs the analysis and hands back the findings as data:

```{r}
result <- arch_check(ui_layer, root = project)

result
```

In your own package the root defaults to the working directory, so you would
simply write `arch_check(ui_layer)`.

## Reading a failure

The violations table has one row per finding:

```{r}
arch_violations(result)
```

Each column answers one question:

| Column | What it tells you |
|---|---|
| `rule` | Which agreement was broken. Matters when you check a rule set. |
| `file`, `line`, `column` | Where, relative to the project root. Your editor can jump straight to it. |
| `enclosing_function` | The innermost named function containing the call, or `NA` at file top level. |
| `callee` | The symbol as written, so `stats::median` and a bare `median` are distinguishable. |
| `owner` | Who the call was attributed to -- a package, or one of your own modules. |
| `resolved_by` | *How* it was attributed. See below. |
| `internal` | `TRUE` when the call used `:::` to reach a package's internals. |
| `source_text` | The call as it appears in the source, so a CI log is enough to act on. |

`resolved_by` takes one of these values:

- `namespace` -- written as `pkg::fn`.
- `namespace_internal` -- written as `pkg:::fn`, reaching into internals.
- `local` -- resolved to a function your own project defines.
- `export` -- an unqualified call attributed to a declared dependency.
- `base` -- an unqualified call attributed to a base package.
- `ambiguous` -- more than one candidate; only present under
  `on_ambiguous = "fail"`.

To fix a violation you either move the call to the layer that should own it, or
you change the rule because the agreement was wrong. Both are fine; leaving a
suppressed violation behind is not an option the package offers.

## Enforcing it in your test suite

`arch_expect()` turns the same check into a `testthat` expectation:

```{r, eval = FALSE}
test_that("UI must not compute statistics", {
  rule("UI must not compute statistics",
    why = "Presentation stays thin so the domain owns the numbers"
  ) |>
    modules_matching("R/ui_*.R") |>
    must_not_call(packages = "stats") |>
    arch_expect()
})
```

The failure message names the rule, its rationale, and every finding. It fails
on violations, and it also fails when a selector matched **no file at all** -- a
rule that selects nothing would otherwise pass vacuously, which is worse than
having no rule.

## Boundaries inside your own code

`must_not_depend_on()` holds one set of your modules away from another:

```{r}
arch_check(
  rule("domain must not reach infrastructure",
    why = "Domain logic has to be testable without a store"
  ) |>
    modules_matching("R/domain_*.R") |>
    must_not_depend_on(modules = "R/infra_*.R"),
  root = project
)
```

No `::` is required -- the constraint works on ordinary unqualified calls
between your files, and a module calling a function it defines itself is never a
violation.

## Forbidding individual functions

Sometimes the package is fine and one function is not:

```{r}
arch_check(
  rule("nothing here may shell out") |>
    modules_matching("R/*.R") |>
    must_not_call(functions = c("system", "setwd")),
  root = project
)
```

Base packages count as declared dependencies, so a bare `system()` resolves to
`base` and the rule fires. A function your project defines itself always wins
over a package export of the same name, so a local helper named `system` would
not be reported.

## A whole architecture as a rule set

Pass a list and get one combined result. The project is discovered, parsed and
resolved **once**, no matter how many rules you check:

```{r}
architecture <- list(
  rule("UI must not compute statistics") |>
    modules_matching("R/ui_*.R") |>
    must_not_call(packages = "stats"),
  rule("domain must not reach infrastructure") |>
    modules_matching("R/domain_*.R") |>
    must_not_depend_on(modules = "R/infra_*.R")
)

combined <- arch_check(architecture, root = project)
combined$rules
```

Every violation row carries the rule that produced it, so a combined result
stays attributable.

## Projects without a DESCRIPTION

A Shiny app, a plumber API or a folder of scripts works the same way. With no
`DESCRIPTION` present, the declared dependencies come from the `library()` and
`require()` calls in the project's own source instead:

```{r}
app <- file.path(tempdir(), "shinyish")
dir.create(file.path(app, "R"), recursive = TRUE, showWarnings = FALSE)

writeLines(
  c("library(stats)", "", "server <- function(input, output) {", "  NULL", "}"),
  file.path(app, "app.R")
)
writeLines(
  c("mod_table_server <- function(id) {", "  median(id)", "}"),
  file.path(app, "R", "mod_table.R")
)

arch_check(
  rule("modules must not compute statistics") |>
    modules_matching("R/mod_*.R") |>
    must_not_call(packages = "stats"),
  root = app
)
```

The bare `median()` call is attributed to `stats` because `app.R` declares it
with `library(stats)`. Both the `library(pkg)` and the `require("pkg")` forms are
recognised. If a `DESCRIPTION` *is* present it wins, and `library()` calls are
ignored.

## Trusting a passing check

This is the part worth reading twice, because a check that resolved nothing
looks exactly like a clean project.

Attributing a bare `dbGetQuery()` to `DBI` means asking `DBI` what it exports,
which requires **`DBI` to be installed** in the library path running the check.
That is rarely a burden -- your test suite already needs your dependencies -- but
a dependency that cannot be loaded degrades resolution rather than erroring, and
its calls land in the unresolved bucket where they can never be violations.

Every result ends with a coverage line:

```{r}
result$counts
```

- **`resolved` well below `call_sites`** means much of your code was not
  attributed, so the rule proved less than it looks like it did.
- **`AMBIGUOUS`** in the printed output means a symbol had more than one
  candidate package and `aRchtest` refused to guess. Qualify the call with `::`,
  or pass `on_ambiguous = "fail"` to treat ambiguity as a violation.
- **`DEGRADED`** means a declared dependency's exports could not be enumerated.
  Install it and re-run before believing a green result.
- **`PARTIAL`** means a file did not parse and was skipped. The rest of the
  project was still analysed -- which is what makes the tool usable mid-refactor
  -- but the analysis is not complete.

## What this package deliberately does not do

Positive and allowlist constraints (`must_only_call()`), layered-architecture
helpers, dependency cycle detection, dynamic dispatch resolution (S4, R6,
`do.call()`), autofix, visualisation and baselining of existing violations are
all out of scope for now. See the README for the reasoning behind each.
